---
title: "HW 2 pitch"
author: "Oleg Slinin"
date: "October 6, 2016"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction

This is a random number generator.  You can select a distribution and 2 parameters, and see a histogram of the results.

## Parameters

### Distribution settings:
- Parameter 1
- Parameter 2
Parameters have different meanings for various distributions.

### Simulation Settings
- N

### Histogram Settings
- Number of buckets

## Example

```{r , echo = F}
brks=25
par(mfrow=c(1,2))
hist(runif(10000, 0, 3), breaks = brks, col = 'darkgray',  border = 'white', main="uniform[0,3]")
hist(rnorm(10000, 0, 3), breaks = brks, col = 'darkgray',  border = 'white', main="normal(0,3)")
```

etc.

## Distributions

- Uniform
- Normal
- binomial
- lognormal
- $\chi^2$

## Problems


- Limited number of distributions.  
- Parameters of distribution can be more than 2.
- Parameters of distribution are not always clear.


